Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}

### 2021-01-26 main.tex
final stitches

First-order predicate logic is a cornerstone of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer science. Upon
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on their own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
made in proof theory in general and in the area of
automated and interactive theorem provers in
particular, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as a possible 24th
problem for his famous lecture~\cite{hilbert:00} in
1900~\cite{thiele:03}, before proof theory existed as a mathematical
field.

@@

A cograph is \bfit{logical} if every vertex is labelled by either an
atom or variable, and it has at least one atom-labelled vertex. An
atom-labelled vertex is called a \bfit{literal} and a
variable-labelled vertex is called a \bfit{binder}. A binder labelled
with $x$ is called an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest proper strong module containing $b$.
An \bfit{$x$-literal} is a literal whose
atom contains the variable $x$. An $x$-binder \bfit{binds} every
$x$-literal in its scope.  In a logical cograph $\gG$, a binder $b$ is
\bfit{existential} (resp. \bfit{universal}) if, for every other vertex
$v$ in its scope, we have $bv \in \eG$ (resp. $bv \notin \eG$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder
and at least one literal.
%
\begin{definition}
A \bfit{first-order graph} or \bfit{fograph} is a logical cograph with legal binders. The
\bfit{binding graph} of a fograph $\gG$ is the directed graph $\bG=\tuple{\vG,
\set{(b, l) \mid b \text{ binds } l}}$.
\end{definition}
%
\noindent We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:
%% For a formula $A$, we can define its
%% associated fograph $\graphof{A}$ inductively by:\todo{duplicate two sentences above}
\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    %% \multicolumn{4}{c}{
    %%   \quad\graphof{\ttt}=\single\ttt
    %%   \qquad
    %%   \graphof{\fff}=\single\fff
    %%   \qquad
    %% \graphof{a} = \single a \hbox{\hskip1em(for any atom $a$)} } \\[1.1ex]
    \graphof{a} &= \single a \rlap{\hskip1em(for any atom $a$)}  \\[.9ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where we
write $\single\alpha$ for a single-vertex  labelled by $\alpha$.

@@

Figure~\ref{fig:KS1} shows the inference rules for the deep inference
system~$\FOKS$ that we introduce in this paper. It is a slight variation
of the systems presented by Br\"unnler~\cite{brunnler:phd} and
Ralph~\cite{ralph:phd} in their PhD-theses. The main differences are
(i) the explicit presence of the $\mixr$-rule, (ii) a
different choice of how the formula equivalence $\fequ$ is defined,
and (iii) an explicit rule for the equivalence.

@@

We will sometimes employ the
general (non-atomic) version of the contraction rule:
\begin{equation}
  \vlinf{\cD}{}{\Scons{A}}{\Scons{A\vlor A}}  
\end{equation}

@@

\begin{proof}
  This is a straightforward induction on the proof of $\sqn\Gamma$ in
  $\FOMLL$, making a case analysis on the bottommost rule instance. We
  show here only the case of
%%  $\downsmash{\vlinf{\fff}{}{\sqn{\Delta,\fff}}{\sqn\Delta}}$ and
  $\vlinf{\forall}{}{\sqn{\Delta,\forall x.A}}{\sqn{\Delta,A}}$ (all
  other cases are simpler or have been shown before,
  e.g.~\cite{brunnler:phd}): By induction hypothesis, there is a proof
  of $\form\Delta\vlor A$ in $\FOMLS$. We can prefix every line in
  that proof by $\forall x$ and then compose the following derivation:
  \begin{center}\vspace{-2ex}\begin{math}
    \vlderivation{
      \vlin{\fequ}{}{\form\Delta\vlor\forall x.A}{
        \vlde{\FOMLS}{}{\forall x.\form\Delta\vlor A}{
          \vlin{\forall}{}{\forall x.\ttt}{
            \vlhy{\ttt}}}}}
  \end{math}\end{center}
  where we can apply the $\fequ$-rule because $x$ is not free in $\Delta$.
\end{proof}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Rule Permutations}

@@

\begin{proof}
  We first permute all instances of $\wD$ and $\wfaD$ to the bottom of
  the derivation and then permute in a second step the rules $\cD$ and
  $\cfaD$ below $\set{\me,\mfaD,\mexD}$. This involves a tedious but
  straightforward case analysis. However, unlike most other rule
  permutations in this paper this has not been done before in the deep
  inference literature. For this reason, we give the full case
  analysis in Appendix~\ref{app:cw-decomposition}. Note that this
  Lemma is the reason for the presence of the rules $\wfaD$ and
  $\cfaD$, as without them the permutation cases in~\eqref{eq:wfa-cfa}
  could not be resolved.
\end{proof}

@@

First observe that in a derivation in $\FOMLS$
where the conclusion is rectified, every line is also rectified, as
the only rules involving bound variables are $\forall$ and $\exists$
which both remove a binder.
Therefore, we can call such a derivation \bfit{rectified}, and
%% if every line in $\Deri$ is
%% rectified.

%% \begin{lemma}
%%   Let  $\Deri$ be a $\FOMLS$ proof of a formula $A$. Then $\Deri$ is rectified iff $A$ is rectified.
%% \end{lemma}

%% \begin{proof}
%%    The only rules involving bound
%%    variables are $\forall$ and $\exists$ which both remove a binder (and
%%    all occurrences of the variable it binds).
%% \end{proof}

%% Hence,
for a non-rectified  $\FOMLS$ derivation $\Deri$ we can
define its \bfit{rectification} $\rectif\Deri$ inductively, by rectifying each
line, proceeding step-wise
from conclusion to premise.\footnote{As for formulas,
the rectification of a derivation is unique up to renaming of bound
variables.}

@@

\begin{proof}[Proof of Theorem~\ref{thm:MLS1->fonet}]
  From $\Deri$ we construct a derivation $\frameof\Deri$ of $\frameof
  A$ in the propositional fragment of $\FOMLS$, such that
  \hbox{$\fographof{\frameof\Deri}=\frameof{\fographof\Deri}$}.  The
  rules $\aiD,\tttD,\mix$ and $\sw$ are translated trivially, and for
  $\fequ$, it suffices to observe that the frame construction is
  invariant under $\fequ$. Finally, for the rules $\forall$ and
  $\exists$, proceed as follows. Every instance of $\forall$ is
  replaced by the derivation on the right below:\footnote{For better
  readability we omit superfluous parentheses, knowing that we always
  have $\fequ$ incorporating associativity and commutativity of
  $\vlan$ and $\vlor$.} \vadjust{\vskip-2ex}
  \begin{equation*}
    \scalebox{.95}{$
    \vlinf{\forall}{}{\Scons{\forall y_j.\ttt}}{\Scons{\ttt}}
    %\qquad
    \;\leadsto
    %\qquad
    \vlderivation{
      \vlde{\set{\sw,\fequ}}{\DDeri_2}{\Scons{q_{h_1,j}\vlor\cdots\vlor
          q_{h_n,j}\vlor(\dual q_{h_1,j}\vlan\cdots\vlan\dual
          q_{h_n,j}\vlan\ttt)}}{
        \vlde{\set{\aiD,\tttD}}{\DDeri_1}{
          \Scons{(q_{h_1,j}\vlor \dual q_{h_1,j})\vlan\cdots\vlan(q_{h_n,j}\vlor \dual q_{h_n,j})\vlan\ttt}}{
          \vlhy{\ttt}}}}
    $}
  \end{equation*}
  where $h_1,\ldots,h_n$ range over the indices of the existential
  binders dependent on that $y_j$. It is easy to see how $\DDeri_1$ is
  constructed, and for $\DDeri_2$ see,
  e.g. \cite{ATS:esslli2019,gug:str:01,brunnler:tiu:01,dissvonlutz}. Then, every
  occurrence of $\forall y_j.F$ is replaced by
  $q_{h_1,j}\vlor\cdots\vlor q_{h_n,j}\vlor (\dual
  q_{h_1,j}\vlan\cdots\vlan\dual q_{h_n,j}\vlan F)$ in the derivation
  below that $\forall$-instance. Now, observe that all instances of
  the $\exists$-rule introducing $x_i$ dependent on $y_j$ must occur
  below in the derivation (otherwise $\Deri$ would not be rectified).
  Now consider such an instance $\vlinf{\exists}{}{\Scons{\exists
      x_i.B}}{\Scons{B\ssubst {x_i} t}}$.  Its context $\Sconhole$
  must contain all the $\forall y_j$ the $\exists x_i$ depends on,
  such that $B$ is in their scope. Following the translation of the
  $\forall$ rules above, we can therefore translate the $\exists$-rule
  instance by the following derivation
  \begin{equation*}
    \vlder{\set{\sw,\fequ}}{\DDeri_3}{ S_0\cons{S_1\cons{\cdots
          S_{l-1}\cons{S_{l}\cons{\dual q_{i,k_1}\vlan\dual
              q_{i,k_2}\vlan\cdots q_{i,k_l}\vlan B'}}\cdots}}}{
      S_0\cons{\dual q_{i,k_1}\vlan S_1\cons{\dual
          q_{i,k_2}\vlan\cdots S_{l-1}\cons{\dual q_{i,k_l}\vlan
            S_{l}\cons{B'}}\cdots}}}
  \end{equation*}
  where $k_1,\ldots,k_l$ are the indices of the universal binders
  on which that $x_i$ depends, and $B'$ is $B$ in which all predicates
  are replaced by a nullary one (step~3 in the frame construction). The
  derivation $\DDeri_3$ can be constructed in the same way as
  $\DDeri_2$.

@@

\begin{remark}
  There is an alternative path of proving Theorem~\ref{thm:MLS1->fonet}
  by translating $\Deri$ to an $\FOMLL$-proof $\Pi$, observing that
  this process preserves the linking and the dualizer. However, for
  this, we have to extend the construction above to the $\cut$-rule,
  and then show that linking and dualizer of a sequent proof $\Pi$ are
  invariant under cut elimination. This can be done similarly to
  unification nets in~\cite{hughes:unifn}.
\end{remark}

@@

In this section we establish the relation between skew bifibrations and derivations  in
$\set{\wrD,\wfaD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}$. However, 
if a derivation $\Deri$ contains instances of the rules $\cfaD$, $\mfaD$, and
$\mexD$ we can no longer naively define the rectification
$\rectif\Deri$ as in the previous section for $\FOMLS$, as these two
rules cannot be applied if premise and conclusion are rectified. For
this reason we define here rectified versions $\rectif\cfaD$,  $\rectif\mfaD$ and
$\rectif\mexD$, shown below:%\vadjust{\vskip-1ex}
\begin{equation*}
  \scalebox{.9}{$
  \vlinf{\rectif\cfaD}{}{\Scons{\forall x.Ax}}{\Scons{\forall y.\forall x.Ax}}
  \hskip3em
  \begin{array}{c}
    \vlinf{\rectif\mfaD}{}{\Scons{\forall x.(Ax\vlor Bx)}}{\Scons{(\forall y.Ay)\vlor(\forall z.Bz)}}
    \\ \\[-1ex]
    \vlinf{\rectif\mexD}{}{\Scons{\exists x.(Ax\vlor Bx)}}{\Scons{(\exists y.Ay)\vlor(\exists z.Bz)}}
  \end{array}
  $}
\end{equation*}
Here, we use the notation $A\cdot$ for a formula $A$ with occurrences
of a placeholder $\cdot$ for a variable. Then $Ax$ stands for the
results of replacing that placeholder with $x$, and also indicating
that $x$ must not occur in $A\cdot$. Then $\forall x.Ax$ and $\forall
y.Ay$ are the same formula modulo renaming of the bound variable bound
by the outermost $\forall$-quantifier. We also demand that the variables $x$, $y$, and $z$ do
not occur in the context $\Sconhole$.

@@

\begin{lemma}\label{lem:fullinj->w}
  Let $\phi\colon\gG\to\gH$ be full and injective skew bifibration
  such that $\rsubstof\phi$ is the identity substitution, and let $G$
  and $H$ be formulas with $\graphof G=\gG$ and $\graphof H=\gH$. Then
  There is a derivation $\vlder{\set{\wrD,\wfaD,\fequ}}{}{H}{G}$.
\end{lemma}

>>>>

First-order predicate logic is a cornerstone of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer science. Upon
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on their own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
made in proof theory in general and in the area of
automated and interactive theorem provers in
particular, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as a possible 24th
problem~\cite{thiele:03} for his famous lecture in
1900~\cite{hilbert:00}, before proof theory existed as a mathematical
field.

@@

A cograph is \bfit{logical} if every vertex is labelled by either an
atom or variable, and it has at least one atom-labelled vertex. An
atom-labelled vertex is called a \bfit{literal} and a
variable-labelled vertex is called a \bfit{binder}. A binder labelled
with $x$ is called an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest proper strong module containing $b$.
An \bfit{$x$-literal} is a literal whose
atom contains the variable $x$. An $x$-binder \bfit{binds} every
$x$-literal in its scope.  In a logical cograph $\gG$, a binder $b$ is
\bfit{existential} (resp. \bfit{universal}) if, for every other vertex
$v$ in its scope, we have $bv \in \eG$ (resp. $bv \notin \eG$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder
and at least one literal.
%
\begin{definition}
A \bfit{first-order graph} or \bfit{fograph} is a logical cograph in which all binders are legal. The
\bfit{binding graph} of a fograph $\gG$ is the directed graph $\bG=\tuple{\vG,
\set{(b, l) \mid b \text{ binds } l}}$.
\end{definition}
%
\noindent We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:
%% For a formula $A$, we can define its
%% associated fograph $\graphof{A}$ inductively by:\todo{duplicate two sentences above}
\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    %% \multicolumn{4}{c}{
    %%   \quad\graphof{\ttt}=\single\ttt
    %%   \qquad
    %%   \graphof{\fff}=\single\fff
    %%   \qquad
    %% \graphof{a} = \single a \hbox{\hskip1em(for any atom $a$)} } \\[1.1ex]
    \graphof{a} &= \single a \rlap{\hskip1em(for any atom $a$)}  \\[.9ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where we
write $\single\alpha$ for a single-vertex  labelled by $\alpha$.

@@

Figure~\ref{fig:KS1} shows the inference rules for the deep inference
system~$\FOKS$ that we introduce in this paper. It is a slight
variation of the systems presented by Br\"unnler~\cite{brunnler:phd}
and Ralph~\cite{ralph:phd} in their PhD-theses. The main differences
are (i) the explicit presence of the $\mixr$-rule, (ii) a different
choice of how the formula equivalence $\fequ$ is defined, (iii) an
explicit rule for the equivalence, and (iv) new inference rules
$\wfaD$ and $\cfaD$. The reason behind these design choices is to
obtain the correspondence with combinatorial proofs and the full
completeness result.

@@

We will for some results also employ the
general (non-atomic) version of the contraction rule:
\begin{equation}
  \vlinf{\cD}{}{\Scons{A}}{\Scons{A\vlor A}}  
\end{equation}

@@

\begin{proof}
  This is a straightforward induction on the proof of $\sqn\Gamma$ in
  $\FOMLL$, making a case analysis on the bottommost rule instance. We
  show here only the case of
%%  $\downsmash{\vlinf{\fff}{}{\sqn{\Delta,\fff}}{\sqn\Delta}}$ and
  $\vlinf{\forall}{}{\sqn{\Delta,\forall x.A}}{\sqn{\Delta,A}}$ (all
  other cases are simpler or have been shown before,
  e.g.~\cite{brunnler:phd}): By induction hypothesis, there is a proof
  of $\form\Delta\vlor A$ in $\FOMLS$. We can prefix every line in
  that proof by $\forall x$ and then compose the following derivation:
  \begin{center}\vspace{-2ex}\scalebox{.9}{\begin{math}
    \vlderivation{
      \vlin{\fequ}{}{\form\Delta\vlor\forall x.A}{
        \vlde{\FOMLS}{}{\forall x.\form\Delta\vlor A}{
          \vlin{\forall}{}{\forall x.\ttt}{
            \vlhy{\ttt}}}}}
  \end{math}}\end{center}
  where we can apply the $\fequ$-rule because $x$ is not free in $\Delta$.
\end{proof}

@@

\begin{remark}
  Observe that Lemma~\ref{lem:cw-atomic} would also hold with the rules $\wfaD$ and~$\cfaD$ removed.
\end{remark}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Rule Permutations}

@@

\begin{proof}
  We first permute all instances of $\wD$ and $\wfaD$ to the bottom of
  the derivation and then permute in a second step the rules $\cD$ and
  $\cfaD$ below $\set{\me,\mfaD,\mexD}$. This involves a tedious but
  straightforward case analysis. However, unlike most other rule
  permutations in this paper, this has not been done before in the deep
  inference literature. For this reason, we give the full case
  analysis in Appendix~\ref{app:cw-decomposition}. Note that this
  Lemma is the reason for the presence of the rules $\wfaD$ and
  $\cfaD$, as without them the permutation cases in~\eqref{eq:wfa-cfa}
  could not be resolved.
\end{proof}

@@

First observe that in a derivation in $\FOMLS$
where the conclusion is rectified, every line is also rectified, as
the only rules involving bound variables are $\forall$ and $\exists$
which both remove a binder.
Therefore, we can call such a derivation \bfit{rectified}, and
%% if every line in $\Deri$ is
%% rectified.
%
%% \begin{lemma}
%%   Let  $\Deri$ be a $\FOMLS$ proof of a formula $A$. Then $\Deri$ is rectified iff $A$ is rectified.
%% \end{lemma}
%
%% \begin{proof}
%%    The only rules involving bound
%%    variables are $\forall$ and $\exists$ which both remove a binder (and
%%    all occurrences of the variable it binds).
%% \end{proof}
%
%% Hence,
for a non-rectified  $\FOMLS$ derivation $\Deri$ we can
define its \bfit{rectification} $\rectif\Deri$ inductively, by rectifying each
line, proceeding step-wise
from conclusion to premise.\footnote{As for formulas,
the rectification of a derivation is unique up to renaming of bound
variables.}

@@

\begin{proof}[Proof of Theorem~\ref{thm:MLS1->fonet}]
  From $\Deri$ we construct a derivation $\frameof\Deri$ of $\frameof
  A$ in the propositional fragment of $\FOMLS$, such that
  \hbox{$\fographof{\frameof\Deri}=\frameof{\fographof\Deri}$}.  The
  rules $\aiD,\tttD,\mix$ and $\sw$ are translated trivially, and for
  $\fequ$, it suffices to observe that the frame construction is
  invariant under $\fequ$. Finally, for the rules $\forall$ and
  $\exists$, proceed as follows. Every instance of $\forall$ is
  replaced by the derivation on the right below:\footnote{For better
  readability we omit superfluous parentheses, knowing that we always
  have $\fequ$ incorporating associativity and commutativity of
  $\vlan$ and $\vlor$.} \vadjust{\vskip-4ex}
  \begin{equation*}
    \scalebox{.95}{$
    \vlinf{\forall}{}{\Scons{\forall y_j.\ttt}}{\Scons{\ttt}}
    %\qquad
    \;\leadsto
    %\qquad
    \vlderivation{
      \vlde{\set{\sw,\fequ}}{\DDeri_2}{\Scons{q_{h_1,j}\vlor\cdots\vlor
          q_{h_n,j}\vlor(\dual q_{h_1,j}\vlan\cdots\vlan\dual
          q_{h_n,j}\vlan\ttt)}}{
        \vlde{\set{\aiD,\tttD}}{\DDeri_1}{
          \Scons{(q_{h_1,j}\vlor \dual q_{h_1,j})\vlan\cdots\vlan(q_{h_n,j}\vlor \dual q_{h_n,j})\vlan\ttt}}{
          \vlhy{\ttt}}}}
    $}
  \end{equation*}
  where $h_1,\ldots,h_n$ range over the indices of the existential
  binders dependent on that $y_j$. It is easy to see how $\DDeri_1$ is
  constructed. The construction of $\DDeri_2$, using $\sw$ and $\fequ$, is standard, see,
  e.g. \cite{ATS:esslli2019,gug:str:01,brunnler:tiu:01,dissvonlutz}. Then, every
  occurrence of $\forall y_j.F$ is replaced by
  $q_{h_1,j}\vlor\cdots\vlor q_{h_n,j}\vlor (\dual
  q_{h_1,j}\vlan\cdots\vlan\dual q_{h_n,j}\vlan F)$ in the derivation
  below that $\forall$-instance. Now, observe that all instances of
  the $\exists$-rule introducing $x_i$ dependent on $y_j$ must occur
  below in the derivation (otherwise $\Deri$ would not be rectified).
  Now consider such an instance $\vlinf{\exists}{}{\Scons{\exists
      x_i.B}}{\Scons{B\ssubst {x_i} t}}$.  Its context $\Sconhole$
  must contain all the $\forall y_j$ the $\exists x_i$ depends on,
  such that $B$ is in their scope. Following the translation of the
  $\forall$ rules above, we can therefore translate the $\exists$-rule
  instance by the following derivation
  \begin{equation*}
    \vlder{\set{\sw,\fequ}}{\DDeri_3}{ S_0\cons{S_1\cons{\cdots
          S_{l-1}\cons{S_{l}\cons{\dual q_{i,k_1}\vlan\dual
              q_{i,k_2}\vlan\cdots q_{i,k_l}\vlan B'}}\cdots}}}{
      S_0\cons{\dual q_{i,k_1}\vlan S_1\cons{\dual
          q_{i,k_2}\vlan\cdots S_{l-1}\cons{\dual q_{i,k_l}\vlan
            S_{l}\cons{B'}}\cdots}}}
  \end{equation*}
  where $k_1,\ldots,k_l$ are the indices of the universal binders
  on which that $x_i$ depends, and $B'$ is $B$ in which all predicates
  are replaced by a nullary one (step~3 in the frame construction). The
  derivation $\DDeri_3$ can be constructed in the same way as
  $\DDeri_2$.

@@

\begin{remark}
  There is an alternative path of proving Theorem~\ref{thm:MLS1->fonet}
  by translating $\Deri$ to an $\FOMLL$-proof $\Pi$, observing that
  this process preserves the linking and the dualizer. However, for
  this, we have to extend the construction from the previous subsection to the $\cut$-rule,
  and then show that linking and dualizer of a sequent proof $\Pi$ are
  invariant under cut elimination. This can be done similarly to
  unification nets in~\cite{hughes:unifn}.
\end{remark}

@@

In this section we establish the relation between skew bifibrations and derivations  in
$\set{\wrD,\wfaD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}$. However, 
if a derivation $\Deri$ contains instances of the rules $\cfaD$, $\mfaD$, and
$\mexD$ we can no longer naively define the rectification
$\rectif\Deri$ as in the previous section for $\FOMLS$, as these two
rules cannot be applied if premise and conclusion are rectified. For
this reason we define here rectified versions $\rectif\cfaD$,  $\rectif\mfaD$ and
$\rectif\mexD$, shown below:\vadjust{\vskip-1ex}
\begin{equation*}
  \scalebox{1}{$
  \vlinf{\rectif\cfaD}{}{\Scons{\forall x.Ax}}{\Scons{\forall y.\forall x.Ax}}
  \hskip3em
  \begin{array}{c}
    \vlinf{\rectif\mfaD}{}{\Scons{\forall x.(Ax\vlor Bx)}}{\Scons{(\forall y.Ay)\vlor(\forall z.Bz)}}
    \\ \\[-1ex]
    \vlinf{\rectif\mexD}{}{\Scons{\exists x.(Ax\vlor Bx)}}{\Scons{(\exists y.Ay)\vlor(\exists z.Bz)}}
  \end{array}
  $}
\end{equation*}
Here, we use the notation $A\cdot$ for a formula $A$ with occurrences
of a placeholder $\cdot$ for a variable. Then $Ax$ stands for the
results of replacing that placeholder with $x$, and also indicating
that $x$ must not occur in $A\cdot$. Then $\forall x.Ax$ and $\forall
y.Ay$ are the same formula modulo renaming of the bound variable bound
by the outermost $\forall$-quantifier. We also demand that the variables $x$, $y$, and $z$ do
not occur in the context $\Sconhole$.

@@

\begin{lemma}\label{lem:fullinj->w}
  Let $\phi\colon\gG\to\gH$ be full and injective skew bifibration
  such that $\rsubstof\phi$ is the identity substitution, and let $G$
  and $H$ be formulas with $\graphof G=\gG$ and $\graphof H=\gH$. Then
  there is a derivation $\vlder{\set{\wrD,\wfaD,\fequ}}{\Deri}{H}{G}$.
\end{lemma}

\begin{proof}
   By \cite[Proposition~7.6.1]{str:07:RTA}, we have a derivation
   $\vlder{\set{\wrD,\PE\fequ}}{\DDeri}{\PE H}{\PE G}$. In order to
   lift $\DDeri$, we need to reorganize the instances of
   $\wD$. If $H$ contains a subformula $\forall x.A$ which is not
   present in $G$, the $\wD$-instances in $\DDeri$ could introduce the
   parts of the propositional encoding $x\vlor A$ independently.  We
   say that an instance $\rr_1$ of $\wD$ in $\Deri$ is \emph{in the
   scope} of an instance $\rr_2$ of $\wD$ if $\rr_1$ introduced
   formulas that contain a free variable $x$ (i.e., $x$ occurs in a
   term in a predicate) and $\rr_2$ introduces the atom $x$ as a
   subformula (i.e. the propositional encoding of the binder $x$).  We
   can now permute the $\wD$-instances in $\DDeri$ such that whenever
   a rule instance $\rr_1$ is in the scope on an instance $\rr_2$, then
   $\rr_2$ occurs below $\rr_1$ in $\DDeri$.
   Then we can lift $\DDeri$ stepwise. First, observe that each line of
   $\DDeri$ is $\PE\fequ$-equivalent to the propositional encoding $\PE P$ of a
   first-order formula $P$.  We now have to show
   that each instance of $\wrD$ in $\DDeri$ is indeed the image of a correct
   application of $\wD$ or $\wfaD$ in first-order logic.
   If we have a $\wD$ of the form
   \begin{equation*}
     \scalebox{1}{$
     \vlinf{\wD}{}{\PE S\cons{x\vlor \PE A}}{\PE S\cons{\PE A}}
     \qquor
     \vlinf{\wD}{}{\PE S\cons{(x\vlor \PE B)\vlor \PE A}}{\PE S\cons{\PE A}}
     $}
   \end{equation*}
   then $x$ cannot occur freely in $A$, as otherwise the fibration
   property would be violated. We can therefore lift these instances to
   \begin{equation*}
      \scalebox{1}{$
    \vlinf{\wfaD}{}{S\cons{\forall x. A}}{S\cons{A}}
     \qquor
     \vlinf{\wD}{}{S\cons{(\forall x. B)\vlor A}}{S\cons{A}}
     $}
   \end{equation*}
   respectively.     
  %% A subformula of $\PE B$ is called \emph{weak} if it has been
  %% introduced by a weakening. All subformulas of a weak formula are
  %% also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  %% \emph{weak pair} if $\PE B\fequ\Scons{B'\vlor B''}$ for some context
  %% $\Sconhole$. We can assume without loss of generality that
  %% whenever weak subformulas $B'$ and $B''$ form a weak pair, they have
  %% been introduced by the same instance of $\wrD$ in
  %% $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  %% brought into this form by simple rule permutations, and then
  %% collapsing several weakenings into a single one.}  Now we show that
  %% $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  %% weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  %% delete all atoms in the scope of the corresponding quantifier,
  %% because $\phi'$ is a fibration on the binding graph. Hence, each line
  If a  weakening happens inside a subformula $x\vlor \PE C$ or $x\vlan
  \PE C$ in $\DDeri$, then there are the following cases:
  \begin{equation*}%\small
    \scalebox{.9}{$
    \vlinf{\wD}{}{\PE S\cons{x\vlor\PE  D\vlor \PE C}}{\PE S\cons{x\vlor\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{x\vlan (\PE D\vlor \PE C)}}{\PE S\cons{x\vlan\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{(x\vlor\PE  D)\vlan \PE C}}{\PE S\cons{x\vlan \PE C}}
    $}
  \end{equation*}
  The first two cases can be lifted to
  \begin{equation*}
    \vlinf{\wD}{}{S\cons{\forall x.( D\vlor C)}}{S\cons{\forall x.  C}}
    \quand
     \vlinf{\wD}{}{S\cons{\exists x.( D\vlor C)}}{S\cons{\exists x.  C}}
  \end{equation*}
  respectively. But in the third case,  an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur. All other situations can be lifted trivially, giving us
  $\vlder{\set{\wrD,\wfaD,\fequ}}{\Deri}{H}{G}$ as desired.
\end{proof}
